Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2021-12-02 sports_analytics.Rmd
added somew football references.

Mathematically, we define $X$ to be a random variable indicating the number of runs (points) that a team will score in the remainder of a period, or game. 
Let $s$ be a tuple that encodes the *state* of a game. 
Then our task is to estimate:
$$
  \mathbb{E}[X | s] \,,
$$
for any state $s$. 
In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* and it is foundational in baseball analytics. 
See [blah] for examples of analytical findings that are based on the expected run matrix.

However, this concept is easily extended to other sports. 
In American football, $s$ is determined by down, yardage to the next first down, and field position. 
Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.

@@

```{r chess-startup, fig.cap="The starting chess board printed via the \\pkg{chess} package."}
library(chess)
Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python")
game() %>% plot()
```

>>>>

Mathematically, we define $X$ to be a random variable indicating the number of runs (points) that a team will score in the remainder of a period, or game. 
Let $s$ be a tuple that encodes the *state* of a game. 
Then our task is to estimate:
$$
  \mathbb{E}[X | s] \,,
$$
for any state $s$.

In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there $24 = 8 \cdot 3$ basic states of an inning in baseball.

The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* and it is foundational in baseball analytics.

See [blah] for examples of analytical findings that are based on the expected run matrix.

However, this concept is easily extended to other sports. 
In American football, $s$ is determined by down, yardage to the next first down, and field position.

Numerous papers (see @lopez2020bigger for details) have used the analysis of the expected number of points to improve fourth down strategy.

@Pattani2012 describes ESPN's expected points metric.

@CarrollPalmer1998 The hidden game of football.  Mentioned by Pattani and popularizing the idea of EPA.

@nflwar, which uses an expected points model as part of the nfl war framework, uses down (1-4, seconds remaining in half, yards from endzone, yards to go (log), indicator for goal to go (GTG), and indicator for under 2 minutes (UTM).

@CarterMachol1971 estimate the expected points for 1st and 10 given a yard line (broken up into 10 yard chunks centered at a given yardline (e.g. 5, 15, 25, 35, etc. )).  This is wild that they did this in 1971!

@@

```{r chess-startup, fig.cap="The starting chess board printed via the \\pkg{chess} package."}
library(chess)
#Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
game() %>% plot()
```
